Supplementary Material

### 2025-07-23 point-by-point-response.qmd

> “We found that the ATLS ® and PTC arms had lower 30-day mortality compared to the PTC and standard care arms. This finding could hint towards a potential effect of training physicians in trauma life support, but it is important to note that this pilot study was not powered to detect any differences in outcomes. The arms differed considerably in sample size, with the ATLS ® arm having the smallest sample size. This difference most likely resulted from the randomisation process with a small number of heterogeneous clusters, and this heterogeneity highlights the importance of taking varying cluster sizes into account in the design of the full scale trial.”

> If this is the case then these outcomes simply should not be reported. Since this is a feasibility trial the major concern should be assessing missing data, which it seems to be, but this statement will be controversial and may distract from the otherwise important and better-supported conclusions. I would suggest removing discussion and reporting of mortality, and instead simply report on whether or not mortality data was missing, and if so was it differentially missing in the different arms of the study. If the study is not powered to detect a difference and no measures of statistical significance were calculated then this is not useful information to the reader.

> The above is the only concern I had when I chose "no" for "Are the discussion and conclusions justified by the results?"

Thank you for pointing this out and we agree that the reporting of the mortality outcomes is problematic for a feasibility trial. We have decided to report mortality outcomes descriptively in the paper, given comments from other reviewers, but revised the discussion to focus on the missing data and to avoid any conclusions about a potential effect of the training.

@@

> Thank you for submitting this relevant paper and the impressive work behind it. It was an honor to review the paper. Here are some suggestions and comments, which I hope will be helpful in further developing the paper.

Thank you for your comments, they were very helpful. We have revised the manuscript in the light of your comments.

> The aim of the study is to gain insight into the feasibility of a larger study. At the same time, more results are presented than necessary to answer this question. This is not clear to the reader.

We agree that the manuscript, and the supplementary material, includes more results than necessary to fulfil the aim of the study. As you have noticed, we struggled to find a balance between adhering to the protocol and reporting the results that were most relevant in a transparent manner. We have now attempted to address this by clearly stating that we deviated from the original protocol in reporting the pilot, to focus on the feasibility of the full-scale trial. We have therefore limited the number of results that we present in the main text of the manuscript considerably.

@@

> In the discussion, deviations from the protocol are mentioned and referred to in the appendix, while these data are actually appropriate to discuss (and share) in the article itself, since it is a pilot.

We have revised the manuscript to describe the deviations where they are most relevant, and retained a list in the supplementary materials for completeness.

> I find the number of appendices in the appendix very extensive, which makes it lose its purpose, and I also wonder to what extent it contributes to answering the research question. Parts of the protocol, which are now included in the appendix, should, in my opinion, be incorporated into the main paper.

We agree that the supplementary materials are very extensive. We included these to transparently report the results of the analyses specified in the original protocol, but in response to your comment we have now published the majority of the supplementary materials as online additional files, and cite these in the manuscript. We have attempted to include parts of the supplementary materials in the main text where appropriate, in response to your earlier comments.

@@

> - For blinding, the authors should state that it was not possible to blind those involved.  It is unclear if those analyzing the data were blinded, as this is possible.

We have added that we did not blind the data analysts in this pilot study, but recommend doing so in the full scale trial.

> - The follow-up period for each participant is unclear. The methods state that data were collected for four months.  The methods also mention a 3-day mortality outcome.

The data collection period in each hospital was intended to be four months, and then patients were followed up for one month. However, the actual data collected periods varied slightly between clusters, and we have clarified this in the methods section.

@@

> - A few key prognostic factors differ across the treatment groups (age, ISS).

Thank you for pointing this out and we now highlight this in the results section.

> - Consider reporting overall enrollment rates (e.g. number of participants enrolled per month).  There is a paragraph in the discussion on low enrollment, but it is not listed in the results.

We have added a figure to the result's section, reporting the number of patients included per month per cluster. Notably, the process of preparing this figure allowed us to identify a data collection error in two clusters, and we have added this to the methods section. We have also included this in the discussion as a limitation of this pilot study and an important lesson for the full scale trial.

@@

> - The EQ-5D should be scored as opposed to showing the results for each question.

Currently, there is no Indian value set available for the EQ-5D-3L, and we have therefore not scored the data. We now explain this in the methods section.

> - Simplify the outcomes tables to focus on the key findings. Remove the yes/no and only list the yes.

We have removed the majority of the outcome tables from the supplementary materials and published them as online additional files, referenced in the manuscript.

@@

> - The absolute difference versus relative difference does not add to the analyses and could be removed.

We have removed the absolute difference versus relative difference from the main results section, but report them in the additional online files as this analysis was part of the original protocol.

@@

> - Comment on the ethics of the full trial. Is it ethical to have sites continue with usual care after seeing the pilot study results?

We have added a sentence on the ethics of the full trial.

> - Consider a stepped wedge design where all clusters start with standard of care and then transition into an intervention.

Thank you for this suggestion and we agree that a stepped wedge design would be a good approach for the full scale trial. We now include this as a suggestion for the full scale trial in the discussion.

@@

> 8. Table S1. Suggest clarifying that this is patient characteristics in the title.

We have added "patient" to the title.

> 9. Table S5. Certain cells are blank – a footnote should explain what this means (no observations?)

Yes, that is correct. Empty cells indicate that there were no observations. We have now moved the majority of the tables from the supplementary materials and instead published them as online additional files, referenced in the manuscript. These files include a  note to clarify that empty cells indicate that there were no observations.

### 2025-07-23 manuscript.Rmd
Compile manuscript

## Trial registration
This pilot study was registered with ClinicalTrials.gov (reg. no NCT05417243). We intended to also register this pilot trial with Clinical Trials Registry - India, but this was not done because of a communication error in the management group.

@@

Previous studies on the effect of ATLS^®^ or PTC training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016]. Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016]. In contrast, some studies have found that these programmes may be associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016]. We found fewer deaths in the intervention arms compared to the standard arm. This difference  may have resulted from the randomisation process with a small number of heterogeneous clusters, highlighting the importance of taking varying cluster sizes into account when designing the full scale trial. A full scale trial remains ethical after this pilot study, considering that it was never powered to detect meaningful differences in clinical outcomes.

Considering the widespread use of trauma life support training, several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020]. Our study represents the first published attempt to pilot a controlled trial of the effect of trauma life support training on patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot study.

>>>>

## Trial registration
This pilot study was registered with ClinicalTrials.gov (reg. no NCT05417243).

@@

Previous studies on the effect of ATLS^®^ or PTC training on patient outcomes are observational or quasi-experimental without a control group, with heterogeneous results [@vestrup_impact_1988, @Ariyanayagam1992; @ali_trauma_1993; @van_olden_clinical_2004; @VanHeng2008; @Wang2010; @drimousis_advanced_2011; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @CioèPeña2016]. Most studies have found that these programmes are associated with improved outcomes, although not all studies have found significant effects [@vestrup_impact_1988; @ali_trauma_1993; @van_olden_clinical_2004; @Wang2010; @noordin_evolution_2011; @hashmi_hospital-based_2013; @Petroze2014; @magnone_impact_2015; @bellanova_mountain_2016; @kamau_impact_2024; @kumar_evaluating_2025]. In contrast, some studies have found that these programmes may be associated with increased mortality [@drimousis_advanced_2011; @CioèPeña2016]. We found fewer deaths in the intervention arms compared to the standard arm. This difference  may have resulted from the randomisation process with a small number of heterogeneous clusters, highlighting the importance of taking varying cluster sizes into account when designing the full scale trial.

A full scale trial remains ethical after this pilot study, considering that it was never powered to detect meaningful differences in clinical outcomes. In addition, educating physicians in trauma life support through programmes like ATLS^®^ and PTC is considered standard care in many settings, but this approach has been criticised for being costly and for propogating outdated practices [@wiles_atls_2015]. Several systematic reviews call for trials in settings where these programmes are not routinely implemented [@Mohammad2013; @Jayaraman2014; @Kadhum2020], and in recognition of their wide use and high face validity, a stepped-wedge design in which all clusters receive the intervention but at randomised time points may be the best option.

Our study represents the first published attempt to pilot a controlled trial of the effect of trauma life support training on patient outcomes, and we conclude that conducting a full-scale cluster randomised trial should be feasible after incorporating the lessons of this pilot study.

### 2025-07-23 supplementary-materials.Rmd
Compile manuscript

| Cluster                                                       | City       | Beds | Type    | Study period | 
+---------------------------------------------------------------+------------+------+---------+--------------+
| Dr R N Cooper Municipal General Hospital                      | Mumbai     | 600  | Public  |              |
| Seth GS Medical College and KEM Hospital                      | Mumbai     | 1800 | Public  |              |
| Lokmanya Tilak Municipal Medical College and General Hospital | Mumbai     | 1850 | Public  |              |
| MEDICA Superspeciality Hospital                               | Kolkata    | 300  | Private |              |
| Medical College                                               | Kolkata    | ??   | Public  |              |
| Nilratan Sircar Medical College & Hospital                    | Kolkata    | 1910 | Public  |              |
| Postgraduate Institute of Medical Education and Research      | Chandigarh | 1400 | Public  |              |
+---------------------------------------------------------------+------------+------+---------+--------------+

>>>>

| Cluster                                                       | City       | Beds | Type    |
+---------------------------------------------------------------+------------+------+---------+
| Dr R N Cooper Municipal General Hospital                      | Mumbai     | 600  | Public  |
| Seth GS Medical College and KEM Hospital                      | Mumbai     | 1800 | Public  |
| Lokmanya Tilak Municipal Medical College and General Hospital | Mumbai     | 1850 | Public  |
| MEDICA Superspeciality Hospital                               | Kolkata    | 300  | Private |
| Medical College                                               | Kolkata    | ??   | Public  |
| Nilratan Sircar Medical College & Hospital                    | Kolkata    | 1910 | Public  |
| Postgraduate Institute of Medical Education and Research      | Chandigarh | 1400 | Public  |
+---------------------------------------------------------------+------------+------+---------+

### 2025-07-23 supplementary-outcome-tables.Rmd
Compile manuscript

>>>>

# Results with 75% confidence intervals (CI)

## Additional Online Table 1. Outcomes in all patients during the entire study period, by treatment arm (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Outcomes in all patients during the entire study period, by treatment arm"]]

## Additional Online Table 2. Outcomes in all patients before training, by treatment arm (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Outcomes in all patients before training, by treatment arm"]]

## Additional Online Table 3. Outcomes in all patients after training, by treatment arm (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Outcomes in all patients after training, by treatment arm"]]

## Additional Online Table 4. Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]

## Additional Online Table 5. Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]

## Additional Online Table 6. Absolute and relative differences in outcomes after training, comparing standard care with ATLS (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]

## Additional Online Table 7. Absolute and relative differences in outcomes after training, comparing standard care with PTC (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]

## Additional Online Table 8. Absolute and relative differences in outcomes after training, comparing ATLS with PTC (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]

## Additional Online Table 9. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS"]]

## Additional Online Table 10. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC"]]

## Additional Online Table 11. Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC (75% CI)

outcome.results.tables[["ci.level.0.75"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC"]]

# Results with 85% confidence intervals (CI)

## Additional Online Table 1. Outcomes in all patients during the entire study period, by treatment arm (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Outcomes in all patients during the entire study period, by treatment arm"]]

## Additional Online Table 2. Outcomes in all patients before training, by treatment arm (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Outcomes in all patients before training, by treatment arm"]]

## Additional Online Table 3. Outcomes in all patients after training, by treatment arm (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Outcomes in all patients after training, by treatment arm"]]

## Additional Online Table 4. Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Absolute change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]

## Additional Online Table 5. Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Relative change from baseline for all outcomes, comparing the period after training with the period before training, by treatment arms"]]

## Additional Online Table 6. Absolute and relative differences in outcomes after training, comparing standard care with ATLS (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in outcomes after training, comparing standard care with ATLS"]]

## Additional Online Table 7. Absolute and relative differences in outcomes after training, comparing standard care with PTC (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in outcomes after training, comparing standard care with PTC"]]

## Additional Online Table 8. Absolute and relative differences in outcomes after training, comparing ATLS with PTC (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in outcomes after training, comparing ATLS with PTC"]]

## Additional Online Table 9. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with ATLS"]]

## Additional Online Table 10. Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing standard care with PTC"]]

## Additional Online Table 11. Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC (85% CI)

outcome.results.tables[["ci.level.0.85"]][["Absolute and relative differences in changes from baseline for all outcomes, comparing ATLS with PTC"]]
